Language Skills in an Ethnically Segmented Labour market: Estonia 1989 -- 2012

### 2018-07-22 paper/ijm.tex
integration monitor regression code file found and added here

This section briefly reviews the institutional and economic background
in Estonia, a country of 1.3M inhabitants,
focusing on the role of ethnic groups and languages.
See \citet{Leping2008} and
\citet{lindemann+saar2011Russian2ndGeneration} for more information.

Before the Second World War, the country was ethnically rather
homogeneous, populated almost exclusively by ethnic Estonians. After the War,
the Russian-speaking population grew rapidly through migration and
reached 40\% by 1989.  The migrants largely settled in Ida-Virumaa, a
mineral-rich industrial area in North-East of the country, and in the
metropolitan area near the capital city Tallinn, and hence the
Russian-speakers are still very unevenly distributed across the country.

Such an influx of mainly
Russian-speaking population
led to two \emph{de facto} official languages
by 1970s.
The society was largely divided across language, and the division was
also reflected at workplace \citep{leppik+vihalemm2015JofBaltStud}.
Certain sectors in the economy, such as the army, railways and the
merchant fleet
were dominated by
Russian-speaking workers while more local industries,
such as agriculture and
local public administration,
were dominated by ethnic Estonians.
Both communities followed the media in their own respective language, the
universities taught many subjects in Russian and Estonian,
and service sector was largely bilingual.  However, the bilingualism
was mainly one-sided as the bulk of Estonians spoke Russian, but not
the other way around.  A particular institution in parts of the Soviet
Union was segregated school system, where Russian schools existed along the
schools teaching in the local language.  Hence a common education
system, one of the central integration mechanisms in the developed world,
was (and still largely is) missing in Estonia.
The widening use of Russian also
caused increasing concerns about the future of the Estonian
language, in turn leading to  unwillingness to use Russian by
the ethnic Estonians.  This further strengthened the language-based segregation.

When the central Soviet authority weakened during the late 1980, the
Estonian-speaking community was quick to organize and establish a
nation-state in 1991 where
Estonian was granted status as the sole official language.  The
above-mentioned ``one-sided bilingualism'' slowly started turning
around.  The universities quickly moved to Estonian-only curriculum,
Estonian language become a large-scale compulsory subject in Russian
schools while teaching Russian in Estonian-language schools declined.
Since 2007, all Russian high schools have to teach part of the
curriculum in Estonian \citep{lindemann+saar2011Russian2ndGeneration}.

According to the report of the Ministry of Education and Research the number
of students, whose mother tongue is not Estonian in basic schools who study in the Estonian language increased from 17.5\% in 2005 to 25.8\% in 2015 due to the growing popularity of immersion
classes (enrolment of classes rose from 7.8\% in 2005 to 17.5\% in 2015).
At the same time the number of students whose mother tongue is not
Estonian studying in Estonian basic schools slowly dropped
respectively from 9.5\% to  8.3\% \citep{HTM2015}.  The younger generation of ethnic Estonians is
increasingly less fluent in Russian while the ethnic Russians are
steadily getting better in Estonian.

The Estonian economy, a part of the Soviet economy up to 1991, successfully
re-oriented to the Western markets and has been growing rapidly since
the mid-1990s.  
\Cref{fig:gdp_growth} indicates that the Estonian economy grew around 7\% per year before the financial crisis in 2008 and after that grew at a lower rate around 3\% per year.
%\fxwarning{ot: a few numbers about economic growth would be good}
%\fxwarning{SKB: I added rather a figure. Pure numbers or table might not give the best impression of the growth.}
\begin{figure}[h]
	\centering                      
	\caption{GDP growth per year for Estonia}\label{fig:gdp_growth}
	\includegraphics[width=0.6\linewidth]{gdp_growth_per_year.eps}
	\caption*{Source: Worldbank - World Development Indicators, Series  NY.GDP.MKTP.KD.ZG}
\fxnote{ot: remove the cryptic series?}
%\fxnote{skb: removed the series from y-axis label. Should I remove the rest as well since it is already stated in the figure caption?}
\end{figure}
 As a bi-product of the re-adjustment, English skills become extremely
valuable while the supply was still rather limited during the early
years of transition.

The economic and political reforms led to substantial
disparities in income and education opportunities
in the early 1990s \citep{Leping2008,
	lindemann+saar2011Russian2ndGeneration}.  The inter-ethnic contacts
are largely limited to superficial encounters at work and in public
space, the number of close private ties remain limited
\citep{korts2009JofBaltStud}.  The
separate worlds are also reflected in media which may present quite
different and occasionally antagonistic viewpoints depending on the
language \citep{Korts2002}.  Around 25\% of Estonians show little
tolerance for Russian-speaking population (the corresponding figure
for Russians is substantially smaller).  A major issue for this group
is Russians' ability and willingness to communicate in Estonian.
\fxwarning{ot: citation here?}
\fxwarning{SKB: Not found yet, otherwise drop it.}

>>>>

This section briefly reviews the institutional and economic background
in Estonia, a country of 1.3M inhabitants,
focusing on the role of ethnic groups and languages.
See \citet{Leping2008} and
\citet{lindemann+saar2011Russian2ndGeneration} for more detailed information.

Before the Second World War, the country was ethnically rather
homogeneous with 90 percent of the population being ethnic Estonians. After the War,
the Russian-speaking population grew rapidly through migration and
reached 40\% by 1989.  The migrants largely settled in Ida-Virumaa, a
mineral-rich industrial area in North-East of the country, and in the
metropolitan area near the capital city Tallinn.  The
Russian-speakers are still very unevenly distributed.

Such an influx of mainly
Russian-speaking population
led to two \emph{de facto} official languages
by 1970s.
The society was largely divided across language, and the division was
also reflected at workplace \citep{leppik+vihalemm2015JofBaltStud}.
Certain sectors in the economy, such as the armed forces, railways and the
merchant fleet
were dominated by
Russian-speaking workers while more local industries,
such as agriculture and
public administration,
were dominated by ethnic Estonians.
Both communities followed the media in their own respective language, the
universities taught many subjects in both Russian and Estonian,
and service sector was largely bilingual.  However, the bilingualism
was mainly one-sided as the bulk of Estonians spoke Russian, but not
the other way around.  A particular institution in parts of the Soviet
Union was segregated school system, where Russian schools existed along the
schools teaching in the local language.  Hence a common education
system, one of the central integration mechanisms in the developed world,
was (and still largely is) missing in Estonia.
The widening use of Russian also
caused increasing concerns about the future of the Estonian
language, in turn leading to  unwillingness to use Russian by
the ethnic Estonians.  This further strengthened the language-based segregation.

When the central Soviet authority weakened during the late 1980, the
Estonian-speaking community was quick to organize, and established a
nation-state in 1991 where
Estonian was granted status as the sole official language.  The
above-mentioned ``one-sided bilingualism'' started turning
around.  The universities quickly moved to Estonian-only curriculum,
Estonian language become a compulsory subject in Russian schools
while teaching Russian in Estonian-language schools declined.
These trends have continued till this day today, according to
\citet{HTM2015}, one quarter of the students from non
Estonian-speaking families studied in Estonian schools.
As a result, the younger generation of ethnic Estonians is
increasingly less fluent in Russian while the ethnic Russians are
steadily getting better in Estonian.

The Estonian economy, a part of the Soviet economy up to 1991, successfully
re-oriented to the Western markets and has been growing rapidly since
the mid-1990s.  
\Cref{fig:gdp_growth} shows that grew averaged at approximately 7\%
annually
before the financial crisis in 2008, afterwards the figure has been around 3\%.
As a bi-product of the re-adjustment, English skills become extremely
valuable while the supply was still rather limited during the early
years of transition.

\begin{figure}[h]
  \centering                      
  \caption{Annual GDP growth in Estonia}\label{fig:gdp_growth}
  \includegraphics[width=0.6\linewidth]{gdp_growth_per_year.eps}
  \caption*{Source: World Bank - World Development Indicators}
\end{figure}

The economic and political reforms led to substantial
disparities in income and education opportunities
in the early 1990s \citep{Leping2008,
	lindemann+saar2011Russian2ndGeneration}.  The inter-ethnic contacts
are largely limited to superficial encounters at work and in public
space, the number of close private ties remain limited
\citep{korts2009JofBaltStud}.  The
separate worlds are also reflected in media which may present quite
different and occasionally antagonistic viewpoints depending on the
language \citep{Korts2002}.
